Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-03-15 ThesisJP/Background/background.tex
Added comment on IO.
Writing section on adaptive beamforming.

If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence energy is distributed or sensed equally in all directions. When multiple omnidirectional elements are arranged in an array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse in space integrated across time.} of ultrasound by diffraction as the size of the array  (the \textit{aperture}) grows. The 3 dB angular width of this beam, and also the angular resolution of the array, is given by \cite{AngelUltrasound}:
\begin{align}\label{eq:res}
\delta\theta_a = \frac{\lambda}{D} = \frac{c}{fD},
\end{align}
\nomi{$\delta\theta_a$}{Angular resolution of an aperture}where \nom{$\lambda$}{Wavelength} and \nom{$f$}{Frequency} is the ultrasound wavelength and frequency respectively, $c$ is the speed of sound, and \nom{$D$}{Aperture size} is the aperture width. With an average speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$, the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree, which at 8 cm range gives a lateral resolution of around 2 mm. The axial resolution is inverse proportional to the pulse length, and is typically better than the lateral resolution.

@@

If time delays are applied on the signal sent to the different elements it is possible to steer and focus the ultrasound beam in space (Fig.\,\ref{fig:das}). This technique is known as \textit{beamforming}, and for ultrasound imaging it is traditionally applied both when transmitting and receiving ultrasound pulses. For medical ultrasound imaging there are two main types of ultrasound probes, linear and phased probes \cite{AngelUltrasound}. A cardiac probe is a phased probe, which means that the image is constructed from beams, or A-mode images, which are gradually steered from $-\theta/2$ to $\theta/2$ radians, resulting in an imaging sector which are $\theta$ radians wide. This kind of image, where A-mode images are acquired from multiple spatial locations in the $xz$-plane, is known as a B-mode image (B for brightness). If A-mode images are steered in the $yz$-plane as well we get what is known as a volumetric B-mode or 3D ultrasound image.

@@

The required axial sampling is dictated by the transmit frequency and the frequency bandwidth of the probe. From Fourier theory we know that a short impulse consist of many frequency components. A short ultrasound pulse, which would yield high axial resolution, therefore requires a large frequency bandwidth. A ultrasound probe has normally a relative bandwidth between 50 and 100 \%. This corresponds to transmitting a pulse which are between 2 and 1 oscillation long. In the front-end of an ultrasound imaging system a high sampling frequency is typically used (e.g. 40 MHz). However, the ultrasound signal is band limited, and the data rate can therefore be reduced in a process known as \textit{\nom{IQ}{In-phase quadrature (ultrasound data format).} demodulation}. In this process the axial sampling rate is reduced to the bandwidth of the imaging system in the following three steps: demodulation with a complex exponential, low-pass filtering, and decimation. The number of samples in one A-mode image is after this process given by:
\begin{align}
N = \frac{2r}{c}B,
\end{align}
\nomi{$N$}{Number of samples in range}where \nom{$B$}{Frequency bandwidth} is the bandwidth, $B = B_wf_c$ (where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth and \nom{$f_c$}{Center frequency of ultrasound pulse} is the center frequency of the transmitted pulse). The total number of samples in an B-mode image is then 
\begin{align}
N_{\text{B-mode}} = N_\theta N.
\end{align}
If our cardiac probe has a relative bandwidth of $80$ \% we have to process almost $50000$ $M$-long vectors of element data every $1/55$ second. The formula for number of FLOPS needed for DAS beamforming is then given as:
\begin{align}
\text{FLOPS}_{\text{DAS}} = MN_{\text{B-mode}}f_{\text{FPS}}
\end{align}
If our cardiac probe has $M=128$ elements, DAS beamforming requires a processing throughput of around $350$ MFLOPS. Hence, far below the peak processing throughput of modern GPUs and CPUs.

@@

\subsection{Adaptive Beamforming}\label{sec:adaptbf}

\begin{figure}[t!]
\subfigure[Delay-and-sum]{
	\includegraphics[width=0.47\textwidth]{img/scenario_das_resp2.png}
}
\subfigure[Capon]{
	\includegraphics[width=0.47\textwidth]{img/scenario_mv_resp2.png}
}
\caption{Array beam pattern with uniform and Capon weights. An interfering source is located at 20\degree. Notice the reduced sidelobe level for the Capon beam pattern in the direction of the interferer.}
\end{figure}

Capon beamforming\footnote{The name ''Capon beamformer'' is due to work by J. Capon \todo{add citation} on seismic arrays \todo{Move to background.}} or minimum variance beamforming.

@@

Not phase aberration correction.

Add section about the computationally complexity. How many flops are required per rx-beam etc...

>>>>

If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence energy is distributed or sensed equally in all directions. When multiple omnidirectional elements are arranged in a linear array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse in space integrated across time.} of ultrasound by diffraction as the size of the array  (the \textit{aperture}) grows. The 3 dB angular width of this beam, and also the angular resolution of the array, is given by \cite{AngelUltrasound}:
\begin{align}\label{eq:res}
\delta\theta_a = \frac{\lambda}{D} = \frac{c}{fD},
\end{align}
\nomi{$\delta\theta_a$}{Angular resolution of an aperture}where \nom{$\lambda$}{Wavelength} and \nom{$f$}{Frequency} is the ultrasound wavelength and frequency respectively, $c$ is the speed of sound, and \nom{$D$}{Aperture size} is the aperture width. With an average speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$, the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree, which at 8 cm range gives a lateral resolution of around 2 mm. The axial resolution is inverse proportional to the pulse length, and is typically better than the lateral resolution.

@@

If time delays are applied on the signal sent to the different elements it is possible to steer and focus the ultrasound beam in space (Fig.\,\ref{fig:das}). This technique is known as \textit{beamforming}, and for ultrasound imaging it is traditionally applied both when transmitting and receiving ultrasound pulses. For medical ultrasound imaging there are two main types of ultrasound probes, linear and phased probes \cite{AngelUltrasound}. A cardiac probe is a phased probe, which means that the image is constructed from beams, or A-mode images, which are gradually steered from $-\theta/2$ to $\theta/2$ radians, resulting in an imaging sector which are $\theta$ radians wide. This kind of image, where A-mode images are acquired from multiple spatial locations in the $xz$-plane, is known as a B-mode image (B for brightness). If A-mode images are acquired from the $yz$-plane as well we get what is known as a volumetric B-mode or 3D ultrasound image. A probe for 3D imaging has several thousand elements arranged as a matrix. In this thesis we will focus on linear arrays for 2D imaging when the topic is adaptive beamforming, and 3D imaging when the topic is adaptive visualization.

@@

The required axial sampling is dictated by the transmit frequency and the frequency bandwidth of the probe. From Fourier theory we know that a short impulse consist of many frequency components. A short ultrasound pulse, which would yield high axial resolution, therefore requires a large frequency bandwidth. A ultrasound probe has normally a relative bandwidth between 50 and 100 \%. This corresponds to transmitting a pulse which are between 2 and 1 oscillation long. In the front-end of an ultrasound imaging system a high sampling frequency is typically used (e.g. 40 MHz). However, the ultrasound signal is band limited, and the data rate can therefore be reduced in a process known as \textit{\nom{IQ}{In-phase quadrature (ultrasound data format).} demodulation}. In this process the axial sampling rate is reduced to the bandwidth of the imaging system in the following three steps: demodulation with a complex exponential, low-pass filtering, and decimation. The number of samples in one A-mode image is after this process given by:
\begin{align}
N = \frac{2r}{c}B,
\end{align}
\nomi{$N$}{Number of samples in range}where \nom{$B$}{Frequency bandwidth} is the bandwidth, $B = B_wf_c$ (where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth and \nom{$f_c$}{Center frequency of ultrasound pulse} is the center frequency of the transmitted pulse). The total number of samples in an B-mode image is then 
\begin{align}
N_{\text{B-mode}} = N_\theta N.
\end{align}
If our cardiac probe has a relative bandwidth of $80$ \% we have to process almost $50000$ $M$-long vectors of element data every $1/55$ second. The formula for the number of FLOPS required for real-time DAS beamforming of cardiac ultrasound is then given as:
\begin{align}
\text{FLOPS}_{\text{DAS}} = MN_{\text{B-mode}}f_{\text{FPS}} = MB_wf_c
\end{align}
If our cardiac probe has $M=128$ elements, DAS beamforming requires a processing throughput of around $300$ MFLOPS. Hence, far below the peak processing throughput of modern GPUs and CPUs. The main challange with DAS beamforming is the amount of data that has to be moved. If one element sample is two bytes long, for our example, around 600 MB has to be moved on to the GPU every second. This is however less than the maximum bandwidth of PCI Express 2.0 (8 GB/s), even if the relative bandwidth, transmit frequency and number of array elements are increased with a total factor of 10.

@@

\subsection{Adaptive beamforming}\label{sec:adaptbf}
Adaptive beamforming has several meanings in the medical ultrasound literature. It can for instance refer to aberration correction which is the task of applying time delays to compensate for the distortion introduced by fatty layers in the human body. However, in this thesis adaptive beamforming will refer to data-dependent apodization (or weighting) derived from a given optimization criterion. The optimization criterion which we focus on in this thesis is the minimum variance distortionless response criterion, which is also known as Capon's method \cite{Capon1969}. A beamformer which uses weights derived using Capon's method will in this thesis mainly be referred to as the \textit{Capon beamformer}. However, it is also known as the minimum variance  (\nom{MV}{Minimum variance beamformer}) beamformer or the minimum variance distortionless respons beamformer (\nom{MVDR}{Minimum variance distortionless respons beamformer (Capon)}).

\begin{figure}[t!]
\subfigure[Uniform weights]{
	\includegraphics[width=0.47\textwidth]{img/scenario_das_resp2.png}\label{fig:das_weights}
}
\subfigure[Capon weights]{
	\includegraphics[width=0.47\textwidth]{img/scenario_mv_resp2.png}\label{fig:capon_weights}
}
\caption{Examples of array beampatterns where the array weights are either uniform (a) or derived using Capon's method (b). An interfering source is located at 20\degree. Notice the reduced sidelobe level for the Capon beampattern in the direction of the interfering source. Image courtesy of Jo Inge Buskenes.}
\label{fig:weights}
\end{figure}

The Capon beamformer produces a weight vector, as introduced in (\ref{eq:das}), based on the impinging signal and an optimization criterion. The problem is as follow \cite{Capon1969}:
\begin{align}
&\min_{\vec{w}} E\{|z[n]|^2\} \rightarrow \min_{\vec{w}} \vec{w}^H \mat{\hat{R}} \vec{w} \label{eq:capon_optimization_criteria} \\
&\text{subjected to } \vec{w}^H\vec{a} = 1,
\end{align}
where \nom{$\mat{\hat{R}}$}{Sample covariance matrix (with regularization)} is a sample covariance matrix. Hence, the resulting weight vector minimizes the output power (the criterion) while maintaining unit gain in the steering direction $\vec{a}$ (distortionless response). The benefit of adaptive beamforming is clearly seen from Fig\,\ref{fig:weights} which shows a sensitivity map (an \textit{array beampattern}) for DAS beamforming with uniform weights and the Capon beamformer. In this simulated case an interfering source is located at 20\degree{} which for the DAS beamformer with uniform weights (\ref{fig:das_weights}) will leak into the beamformer output with its power attenuated with around $13$ dB. Since the Capon weights are data-dependent the method "knows" where the interfering source is located and is able to position a zero in the direction of it. Hence, the interfering source will be highly attenuated independent of its direction. Another interesting case is when the interfering source is located within the mainlobe of the array beampattern. In theory a signal will be regarded by the Capon beamformer as an interfering source as long as the signal propagation vector does not match the steering vector ($\vec{a}$) exactly. If it is not possible to move a zero to the direction of the interferer the mainlobe will get shifted if this results in minimum output power. Adaptive positioning of zeros and mainlobe shifting gives rice to the super-resolution capabilities of the Capon beamformer \cite{Synnevag2007}, which is the main reason why Capon beamforming has been highly studied in the field of medical ultrasound imaging the last decade.

To "know" the direction of interfering sources the Capon beamformer has to estimate the statistics of the impinging wave field. This is done by estimating a sample covariance matrix, which for an active broadband system can be estimated in the following way \cite{Synnevag2009}:
\begin{align}
\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,
\end{align}
where $K = M-L+1$\nomi{$K$}{Temporal averaging parameter (Capon)} typically is proportional to the pulse length, $N_K = 2K + 1$ \nomi{$N_K$}{Number of samples used for temporal averaging (Capon)}, $N_L = M-L+1$,\nomi{$N_L$}{Number of subarrays used for spatial averaging (Capon)}\nomi{$L$}{Subarray length (Capon)} and $\vec{x}_l$ is a $L$ long subarray $[x_l[n], \dotso x_{l+L}[n]]$. To ensure robustness against model errors and numerical stability the matrix is loaded with a diagonal factor, 
\begin{align}
\epsilon &= d*\text{trace}\{\mat{\breve{R}}\}/L\\
\mat{\hat{R}} &= \mat{\breve{R}} + \epsilon\mat{I}.
\end{align} 
Thus, a white noise component proportional to the in-coherent signal power is added to the estimate.

The solution to the minimization problem in (\ref{eq:capon_optimization_criteria}) is, by the method of Lagrange multipliers \cite{VanTrees2003}, found to be
\begin{align}\label{eq:capon_weights}
\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}} \in \mathbb{C}^L.
\end{align}
Since $\vec{w}$ is time (or range) dependent one matrix has to be constructed and inverted for each data vector $\vec{x}[n]$ received by the system. This is indeed computationally demanding, something which becomes clear when we take a closer look at each step of the beamformer. Construction of the covariance matrix has a computationally complexity of $O(L^2N_LN_K)$\nomi{O(.)}{Big O notation (algorithm complexity)}, inverting it has a computationally complexity of $O(L^3)$, and the final normalization and calculation of $\vec{w}$ is $O(L)$. For $L=M/2$, a subarray length often used in the literature, the first two steps both have a complexity of $O(L^3)$. These steps should therefore receive the same amount of attention when a fast implementation of Capon beamforming is attempted. A rough estimate of the number of FLOPS required by a Capon beamformer for real-time cardiac ultrasound imaging is then given by:
\begin{align}
\text{FLOPS}_{\text{Capon}} = L^3N_{\text{B-mode}}f_{\text{FPS}} = L^3B_wf_c.
\end{align}
Which for our example, with $L=M/2=64$, sums up to a required throughput of more than $0.5$ TFLOPS. Even though this is just 10\% of the peak throughput of modern GPUs remember that the latter where only theoretical numbers, and that the actual throughput is highly algorithm dependent. The required throughput for Capon beamforming is, for the moment, hard to achieve in practice \cite{So2011}. As for DAS beamforming this number also scales with higher relative bandwidth, pulse frequency, and more array elements since subarray size typically is proportional to the number of array elements. In the literature on medical ultrasound imaging we find few attempts on implementing a real-time Capon beamformer for medical ultrasound imaging. In addition to the papers in this thesis only Chen \textit{et al.} \cite{Chen2011, Chen, Chen2011a} are known to have explored GPU and FPGA implementations of the Capon beamformer for medical ultrasound imaging, however with some shortcomings  which are discussed in both \textbf{Paper I} and \textbf{II}.

To reduce the processing requirement for Capon beamforming several simplifications have been proposed \cite{Asl2012, Kim}. The most interesting proposals include the low complexity beamformer by Synnev\aa{}g \textit{et al.} \cite{Synnevag2011} and the beamspace adaptive beamformer by Nilsen and Hafizovic \cite{Nilsen2009}.

Capon beamforming and focused systems.

The Capon beamformer is derived for narrowband, far-field applications and has typically been applied on medical ultrasound imaging "as is" even tough medical ultrasound imaging is a broadband, near-field imaging application. Later in this section we will explain why a broadband version of the Capon beamformer will have little effect. First we will take a closer look at the details of Capon beamforming.

@@
